---
title: "Workshop II: Classification with Logistic Regression"
author: "Roopa Yashaswini Kotha"
date: October 2024
output:
  html_document:
    theme: journal
    highlight: tango
    number_sections: yes
    toc: yes
    toc_float: no
    mathjax: "default"
  pdf_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load_libraries, include = FALSE}
library(tidyverse) # the usual stuff: dplyr, readr, and other goodies
library(lubridate) # to handle dates
library(GGally) # for correlation-scatter plot matrix
library(ggfortify) # to produce residual diagnostic plots
library(rsample) # to split dataframe in training- & testing sets
library(janitor) # clean_names()
library(broom) # use broom:augment() to get tidy table with regression output, residuals, etc
library(huxtable) # to get summary table of all models produced
library(caret) # to train more advanced models (k-fold cross-validation, stepwise regression, LASSO)
library(nnet) # to calculate the maximum value of a vector
library(pROC) # to plot ROC curves
library(sampling)
library(MLmetrics) #for caret LASSO logistic regression
library(data.table)
library(glmnet)

```

# Introduction

## Load the data

First we need to start by loading the data.

```{r, load_data, warning=FALSE, message=FALSE}

lc_raw <- read_csv("LendingClub Data.csv",  skip=1) %>%  
  clean_names() # use janitor::clean_names()
```

# ICE the data: Inspect, Clean, Explore

## Inspect the data

```{r, Inspect}
glimpse(lc_raw)
```

## Clean the data

```{r, clean data}
lc_clean<- lc_raw %>%
  select(-where(~ mean(is.na(.)) > 0.9)) %>% # drop columns with more than 90% missing data
  filter(!is.na(int_rate)) %>%   #delete empty rows
  mutate(
    issue_d = mdy(issue_d),  # lubridate::mdy() to fix date format
    term = factor(term_months),     # turn 'term' into a categorical variable
    delinq_2yrs = factor(delinq_2yrs) # turn 'delinq_2yrs' into a categorical variable
  ) %>% 
  mutate(default = dplyr::recode(loan_status, 
                      "Charged Off" = "1", 
                      "Fully Paid" = "0"))%>%
    mutate(default = as.factor(default)) %>%
  dplyr::select(-emp_title,-installment, -term_months, everything()) #move some not-so-important variables to the end. 
    
```

## Explore the data **(Q1)**

```{r, visualization of defaults, warning=FALSE}
#bar chart of defaults
def_vis1 <- ggplot(data=lc_clean, aes(x=default)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) + 
  labs(x="Default, 1=Yes, 0=No", y="relative frequencies") +
  scale_y_continuous(labels=scales::percent) +
  geom_text(aes( label = scales::percent((..count..)/sum(..count..) ),y=(..count..)/sum(..count..) ), stat= "count",vjust=-0.5) 

def_vis1

#bar chart of defaults per loan grade
def_vis2 <- ggplot(data=lc_clean, aes(x=default), group=grade) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill = factor(..x..)), stat="count")  + 
  labs(title="Defaults by Grade", x="Default, 1=Yes, 0=No", y="relative frequencies")+
  scale_y_continuous(labels=scales::percent) +facet_grid(~grade) + theme(legend.position = "none") +
  geom_text(aes( label = scales::percent((..count..)/sum(..count..) ),y=(..count..)/sum(..count..) ), stat= "count",vjust=-0.5) 
def_vis2

#bar chart of defaults per number of Delinquencies
def_vis3 <- lc_clean %>%
  filter(as.numeric(delinq_2yrs)<4) %>%
  ggplot(aes(x=default), group=delinq_2yrs) +
  geom_bar(aes(y = (..count..)/sum(..count..), fill = factor(..x..)), stat="count")  + 
  labs(title="Defaults by Number of Delinquencies", x="Default, 1=Yes, 0=No", y="relative frequencies")  +
  scale_y_continuous(labels=scales::percent) +facet_grid(~delinq_2yrs) + theme(legend.position = "none") +
  geom_text(aes( label = scales::percent((..count..)/sum(..count..) ),y=(..count..)/sum(..count..) ), stat= "count",vjust=-0.5)

def_vis3

#scatter plots 

#We select 2000 random loans to display only to make the display less busy. 
set.seed(1234)
reduced<-lc_clean[sample(0:nrow(lc_clean), 2000, replace = FALSE),]%>%
  mutate(default=as.numeric(default)-1) # also convert default to a numeric {0,1} to make it easier to plot.

          
# scatter plot of defaults against loan amount  
# If the jitter plot does not show up, try to remove I() and simply use x=loan_amnt/1000
def_vis4 <- ggplot(data=reduced, aes(y=default,x=I(loan_amnt/1000)))  + 
  labs(y="Default, 1=Yes, 0=No", x="Loan Amnt (1000 $)") +
  geom_jitter(width=0, height=0.05, alpha=0.7) #We use jitter to offset the display of defaults/non-defaults to make the data easier to interpert. We have also changed the amount to 1000$ to reduce the number of zeros on the horizontal axis.

def_vis4

#scatter plot of defaults against loan amount.
def_vis5 <- ggplot(data=reduced, aes(y=default,x=I(annual_inc/1000)))   + labs(y="Default, 1=Yes, 0=No", x="Annual Income(1000 $)") +geom_jitter(width=0, height=0.05, alpha=0.7) +  xlim(0,400)

def_vis5

```

We also estimate a correlation table between defaults and other
continuous variables.

```{r, correlation table, warning=FALSE, message=FALSE}

# correlation table using GGally::ggcor()
# this takes a while to plot

lc_clean %>% 
    mutate(default=as.numeric(default)-1)%>%
  select(loan_amnt, dti, annual_inc, default) %>% #keep Y variable last
 ggcorr(method = c("pairwise", "pearson"), label_round=2, label = TRUE)

```

```{r, eval = TRUE}
# My visualization
# Load the data and define interest rate groups
lc_clean %>% 
  mutate(
    # Create a new variable 'int_group' that categorizes 'int_rate' into ranges
    int_group = cut(int_rate, breaks = c(0.05, 0.10, 0.15, 0.20, 0.25), 
                    labels = c("5%-10%", "10%-15%", "15%-20%", "20%-25$%"))
  ) %>% 
  # Start ggplot with 'default' on x-axis, grouping by 'int_group'
  ggplot(aes(x = default), group = int_group) +
  
  # Create a bar plot showing the proportion of defaults within each group
  geom_bar(aes(y = (..count..) / sum(..count..), fill = factor(..x..)), stat = "count")  + 
  
  # Add labels and title for the plot
  labs(
    title = "Defaults by Interest Rate", 
    x = "Default, 1=Yes, 0=No", 
    y = "Relative Frequencies"
  )  +
  
  # Convert y-axis values to percentages for easier interpretation
  scale_y_continuous(labels = scales::percent) +
  
  # Create separate panels (facets) for each 'int_group' category
  facet_grid(~int_group) + 
  
  # Remove the legend for a cleaner look
  theme(legend.position = "none") +
  
  # Display percentage labels on top of each bar
  geom_text(
    aes(label = scales::percent((..count..) / sum(..count..)), 
        y = (..count..) / sum(..count..)), 
    stat = "count",
    vjust = -0.5
  )

```

The above visualisation provides a bar chart on the percentage of
defaults based on the interest rates where the interest rates are
grouped with an interval of 5%. From the above chart, maximum people
were lent loans with interest rate of 10-15% and has the highest
defaults as well.

# Linear vs. logistic regression for binary response variables **(Q2)**

It is certainly possible to use the OLS approach to find the line that
minimizes the sum of squared errors when the dependent variable is
binary (i.e., default / no default). In this case, the predicted values
take the interpretation of a probability. We can also estimate a
logistic regression instead. We do both below.

```{r, linear and logistic regression with binary response variable, warning=FALSE}

model_lm<-lm(as.numeric(default)~I(annual_inc/1000), lc_clean)
summary(model_lm)


logistic1<-glm(default~I(annual_inc/1000), family="binomial", lc_clean)
summary(logistic1)


ggplot(data=reduced, aes(x=annual_inc/1000, y=default)) + 
  geom_smooth(method="lm", se=0, aes(color="OLS"))+ 
  geom_smooth(method = "glm", method.args = list(family = "binomial"),  se=0, aes(color="Logistic"))+ 
  labs(y="Prob of Default", x="Annual Income(1000 $)")+  
  xlim(0,450)+scale_y_continuous(labels=scales::percent)+geom_jitter(width=0, height=0.05, alpha=0.7) + 
  scale_colour_manual(name="Fitted Model", values=c("blue", "red"))

```

Logistic regression is better model because linear regression is useful
for predicting values and can predict probabilities less than 0 or
higher than 1 which is not correct. In the graph above, the linear model
line goes below 0%. However, the logistic regression predicts the
probabilities in correct range, making it easier to classify.

# Multivariate logistic regression **(Q3, Q4)**

```{r, multivariate logistic regression, warning=FALSE, eval = TRUE}
logistic2<- glm(default~annual_inc + term + grade + loan_amnt, family="binomial", lc_clean)
summary(logistic2)

#compare the fit of logistic 1 and logistic 2
anova(logistic1,logistic2)

```

Logistic 2 model is better because it allocates more variables that
reduces the deviance from 31130 to 29286.

```{r, eval = TRUE}
#Predict the probability of default
prob_default<-predict(logistic2,lc_clean,type="response")

#plot 1: Density of predictions
g0<-ggplot( lc_clean, aes( prob_default ) )+
  geom_density( size=1)+
  ggtitle( "Predicted Probability with Logistic 2" )+  
  xlab("Estimated Probability")
g0

#plot 2: Density of predictions by default
g1<-ggplot( lc_clean, aes( prob_default, color=default) ) +
  geom_density( size=1)+
  ggtitle( "Predicted Probability with Logistic 2 by Default" )+  
  xlab("Estimated Probability")
g1
```

## From probability to classification

The logistic regression model gives us a sense of how likely defaults
are; it gives us a probability estimate. To convert this into a
prediction, we need to choose a cutoff probability and classify every
loan with a predicted probability of default above the cutoff as a
prediction of default (and vice versa for loans with a predicted
probability below this cutoff).

Let's choose a threshold of 16%. Of course some of our predictions will
turn out to be right but some will turn out to be wrong -- you can see
this in the density figures of the previous section. Let's call
"default" the "positive" class since this is the class we are trying to
predict. We could be making two types of mistakes. False positives
(*i.e.,* predict that a loan will default when it will not) and false
negatives (*i.e.,* predict that a loan will not default when it does).
These errors are summarized in the confusion matrix.

```{r, From probability to classification, eval = TRUE}
#using the logistic 2 model predict default probabilities
prob_default2<- predict(logistic2,lc_clean,type="response")

#Call any loan with probability more than 16% as default and any loan with lower probability as non-default. Make sure your prediction is a factor with the same levels as the default variable in the lc_clean data frame
p_class<- ifelse(prob_default2 >= 0.16, "1", "0")     

#produce the confusion matrix and set default as the positive outcome
con2<- confusionMatrix(as.factor(p_class), as.factor(lc_clean$default), positive="1")         

#print the confusion matrix
con2

```

> **Q7**. Testing the cutoff to be 20%.

```{r, From probability to classification v2, eval = TRUE}
p_class_updated<- ifelse(prob_default2 >= 0.20, "1", "0") # insert your code

#produce the confusion matrix and set default as the positive outcome
con2_updated<- confusionMatrix(as.factor(p_class_updated), as.factor(lc_clean$default), positive="1") # insert your code
  
#print the confusion matrix
con2_updated

```

20% cutoff has higher accuracy and low sensitivity. However, in the case
of lending club, since predicting defaults is very important, 16% cutoff
which provides higher sensitivity is better.

```{r, ROC curves, warning=FALSE, eval = TRUE}
# Calculate the ROC curve for 'logistic2' model
# 'roc' function takes the actual default outcomes and predicted probabilities of default (prob_default2)
ROC_logistic2 <- roc(lc_clean$default, prob_default2)

# Estimate the Area Under the Curve (AUC) for the 'logistic2' model
# Extract the AUC from the ROC object and store it in 'AUC2'
AUC2 <- ROC_logistic2$auc

# Plot the ROC curve with the ggroc function and display the AUC in the plot title
ggroc(ROC_logistic2, alpha = 0.5) + 
  
  # Add a title to the plot, embedding the AUC value rounded to two decimal places
  ggtitle(paste("Model Logistic 2: AUC =", round(AUC2 * 100, digits = 2), "%"))  +
  
  # Add a dashed diagonal line representing a random classifier (where TPR = FPR)
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed") +
  
  # Add dashed segments for the x and y-axis extremes (TPR = 1 and FPR = 1) for clearer reference
  geom_segment(aes(x = 1, xend = 1, y = 0, yend = 1), color = "black", linetype = "dashed") +
  geom_segment(aes(x = 1, xend = 0, y = 1, yend = 1), color = "black", linetype = "dashed")


```

```{r, out-of-sample ROC curve, eval = TRUE}
# splitting the data into training and testing
set.seed(24)
train_test_split <- initial_split(lc_clean, prop = 0.8)
testing <- testing(train_test_split) #20% of the data is set aside for testing
training <- training(train_test_split) #80% of the data is set aside for training

# run logistic 2 on the training set 
logistic2<- glm(default~annual_inc + term + grade + loan_amnt, family="binomial", training)            # insert your code

#calculate probability of default in the training sample 
p_in<- predict(logistic2, training, type="response")             
  
#ROC curve using in-sample predictions
ROC_logistic2_in <- roc(training$default, p_in)  

#AUC using in-sample predictions
AUC_logistic2_in<- ROC_logistic2_in$auc 
  
#calculate probability of default out of sample 
p_out<- predict(logistic2, testing, type="response")  

#ROC curve using out-of-sample predictions
ROC_logistic2_out <- roc(testing$default, p_out)           
AUC_logistic2_out<- ROC_logistic2_out$auc 
  
#plot in the same figure both ROC curves and print the AUC of both curves in the title
ggroc(list("Logistic 2 In"=ROC_logistic2_in, "Logistic 2 Out"=ROC_logistic2_out))+
  ggtitle(
    paste("Model Logistic 2 In: AUC=",round(AUC_logistic2_in*100, digits=2),
          "%\nModel Logistic 2 Out: AUC=",round(AUC_logistic2_out*100, digits=2),
          "%")
  )+
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color="grey", linetype="dashed")
```

There is no evidence of overfitting because the AUC for both training
and testing sample are nearly same. If the AUC for testing is very less
than training, there could be a chance of over-fitting.

```{r, out-of-sample ROC curve overfitting, eval = TRUE}

#add code here
# Set random seed for reproducibility
set.seed(123)

# Store the number of rows in the data
n <- nrow(lc_clean)

# Generate 500 random predictor variables and add them to the dataset
for (i in 1:500) {
  # Create a new column "random_var_i" for each iteration with random normal values (mean=0, sd=100)
  lc_clean[[paste0("random_var_", i)]] <- rnorm(n, mean = 0, sd = 100)
}

# Set another seed to ensure reproducible train-test splitting
set.seed(24)

# Split data into training (80%) and testing (20%) sets
train_test_split <- initial_split(lc_clean, prop = 0.8)
testing_random <- testing(train_test_split)       # 20% of the data for testing
training_random <- training(train_test_split)     # 80% of the data for training

# Specify the predictors, including original and 500 random variables
predictors <- paste(
  c("annual_inc", "term", "grade", "loan_amnt", paste0("random_var_", 1:500)), 
  collapse = " + "
)

# Create the formula for the logistic regression model
formula <- as.formula(paste("default ~", predictors))

# Train a logistic regression model on the training set
logistic_model <- glm(formula, family = "binomial", data = training_random)

# Make in-sample predictions on the training set
p_random_in <- predict(logistic_model, training_random, type = "response")

# Generate the ROC curve for in-sample predictions
ROC_logistic_random_in <- roc(training_random$default, p_random_in)

# Calculate the AUC for the in-sample ROC curve
AUC_logistic_random_in <- ROC_logistic_random_in$auc

# Make out-of-sample predictions on the testing set
p_random_out <- predict(logistic_model, testing_random, type = "response")

# Generate the ROC curve for out-of-sample predictions
ROC_logistic_random_out <- roc(testing_random$default, p_random_out)

# Calculate the AUC for the out-of-sample ROC curve
AUC_logistic_random_out <- ROC_logistic_random_out$auc

# Plot both ROC curves (in-sample and out-of-sample) on the same plot
ggroc(list("Logistic 2 In" = ROC_logistic_random_in, "Logistic 2 Out" = ROC_logistic_random_out)) +
  
  # Add a title showing the AUC values for both curves
  ggtitle(
    paste(
      "Model Logistic Random In: AUC=", round(AUC_logistic_random_in * 100, digits = 2), "%",
      "\nModel Logistic Random Out: AUC=", round(AUC_logistic_random_out * 100, digits = 2), "%"
    )
  ) +
  
  # Add a diagonal dashed line (random classifier baseline)
  geom_segment(aes(x = 1, xend = 0, y = 0, yend = 1), color = "grey", linetype = "dashed")

```

The difference between AUC sample in and sample out has changed
negatively. The AUC for training is greater than AUC sample out but
there is no significant over-fitting. The reduce in AUC for out sample
could be due to the noise from random variables.

## Selecting loans to invest in using the model Logistic 2.

```{r}
# splitting the data into training and testing
set.seed(24)
train_test_split <- initial_split(lc_clean, prop = 0.6)
training <- training(train_test_split) #60% of the data is set aside for training
remaining <- testing(train_test_split) #40% of the data is set aside for validation & testing
set.seed(2024)
train_test_split <- initial_split(remaining, prop = 0.5)
validation<-training(train_test_split) #50% of the remaining data (20% of total data) will be used for validation
testing<-testing(train_test_split) #50% of the remaining data (20% of total data) will be used for testing
```

```{r, eval = TRUE}
#estimate the model on the training set
# Train a logistic regression model (logistic2) to predict 'default' using specified predictors
logistic2 <- glm(default ~ annual_inc + term + grade + loan_amnt, family = "binomial", data = training)

# Make predictions on the validation set using the logistic2 model
# 'type = "response"' returns predicted probabilities of default
p_val <- predict(logistic2, validation, type = "response")

# Define profit and loss per classification outcome
unit_profit <- 15   # profit per correctly predicted non-default (TN)
unit_loss <- 75     # loss per incorrectly predicted default (FP)

# Initialize vectors to store profit and threshold values
profit <- 0           # profit at each threshold
threshold <- 0        # threshold values (from 0.0025 to 0.25, incremented by 0.0025)

# Loop over 100 threshold values from 0.0025 to 0.25 in steps of 0.0025
for(i in 1:100) {
  # Calculate the current threshold value
  threshold[i] <- i / 400
  
  # Classify predictions as 1 (default) or 0 (non-default) based on the current threshold
  classify_one_or_zero <- ifelse(p_val > threshold[i], 1, 0)
  
  # Generate a confusion matrix to evaluate predictions against actual outcomes
  cm <- confusionMatrix(as.factor(classify_one_or_zero), as.factor(validation$default), positive = "1")
  
  # Calculate profit: true negatives (cm$table[1,1]) yield profit; false positives (cm$table[1,2]) incur loss
  profit[i] <- cm$table[1, 1] * unit_profit - cm$table[1, 2] * unit_loss
}

# Plot the profit curve over different threshold values
ggplot(as.data.frame(threshold), aes(x = threshold, y = profit)) +
  geom_smooth(method = 'loess', se = 0) +
  labs(title = "Profit curve with random classifier")

# Find the maximum profit and corresponding threshold
profit_max <- max(profit)
threshold_max <- threshold[which.max(profit)] * 100

# Print the maximum profit and the optimal threshold as percentages
print(profit_max)
print(threshold_max)

```

```{r}
# Generate predicted probabilities of default for the testing set using the logistic2 model
p_test <- predict(logistic2, testing, type = "response")

# Classify predictions as 1 (default) or 0 (non-default) based on a threshold of 16.5%
classify_one_or_zero_test <- ifelse(p_test > 0.165, 1, 0)

# Create a confusion matrix comparing predictions with actual default values in the testing set
cm_test <- confusionMatrix(as.factor(classify_one_or_zero_test), as.factor(testing$default), positive = "1")

# Calculate profit on the testing set:
# Profit = (True Negatives * unit profit) - (False Positives * unit loss)
profit_test <- cm_test$table[1, 1] * unit_profit - cm_test$table[1, 2] * unit_loss

# Print the profit obtained on the testing set
print(profit_test)

```

Insert your comments here: The threshold probability that generates
maximum profit is **16.5** the corresponding profit generated is
**31050.** The profit for test set is **16500.**

```{r, eval = TRUE}
#add code here
# Classify observations in the validation set as 1 (default) or 0 (non-default) using a 16.5% threshold
classify_mt <- ifelse(p_val > 0.165, 1, 0)

# Initialize counters for missed non-defaults and total return on investment (ROI)
misses <- 0    # Count of defaults that were incorrectly classified as non-defaults
ROI <- 0       # Total ROI based on classifications
counter <- 1   # Initialize a counter for row indexing in the loop

# Loop through each classification result in 'classify_mt'
for (i in classify_mt) {
  
  # Check if the classification is 0 (non-default) and the actual outcome in the validation set is also 0
  if (i == 0 & validation[counter,]$default == 0) {
    
    # If correct non-default prediction, calculate return based on loan terms
    # ROI is increased by ((term * installment amount / loan amount) - 1), which represents loan return
    ROI <- ROI + ((((validation[counter,]$term_months * validation[counter,]$installment) / validation[counter,]$loan_amnt) - 1))
  
  # Check if classification is 0 (non-default) but actual outcome is 1 (default)
  } else if (i == 0 & validation[counter,]$default == 1) {
    
    # If incorrect non-default prediction, apply a fixed loss of 60% of the loan amount
    ROI <- ROI - 0.6
    
    # Increase the 'misses' counter to record the misclassification
    misses <- misses + 1
  }
  
  # Move to the next row in the validation set
  counter <- counter + 1
}

# Calculate the average ROI as a percentage
avg_ROI <- (ROI / nrow(validation)) * 100

# Print the average ROI as a percentage
print(avg_ROI)


```

The average return after investing \$1 on all investments is **11.11%**

```{r, eval = TRUE}
#add code here
# Define a function 'avg_return' that calculates the average return on investment (ROI)
# given a specific cost added to the default scenario
avg_return <- function(cost) {
  
  # For each loan in the validation set, calculate 'loan_return' based on whether there was a default
  # - If no default (default == "0"), return = (installment * term months / loan amount) - 1, indicating profit
  # - If default (default == "1"), return = -0.6 - cost, where -0.6 is the standard loss and 'cost' is an additional fixed cost
  validation <- validation %>%
    mutate(loan_return = ifelse(default == "0", installment * term_months / loan_amnt - 1, -0.6 - cost))
  
  # Calculate the average ROI (avg_ROI) across all loans in the validation set
  avg_ROI <- mean(validation$loan_return)
  
  # Return the average ROI value
  return(avg_ROI)
}

# Define a range of cost values from 0 to 1 in increments of 0.01 to evaluate
cost_range <- seq(0, 1, 0.01)

# Apply the 'avg_return' function across each cost in 'cost_range' using 'sapply'
# - This produces a vector 'avg_returns_values' of average ROI values for each cost in the range
avg_returns_values <- sapply(cost_range, avg_return)

# Identify the maximum cost for which the average ROI is still positive
max_cost <- max(cost_range[avg_returns_values > 0])

# Print the maximum allowable cost that maintains a positive average ROI
print(max_cost)
```

A fixed cost of **\$0.74** for every missed prediction of default would
still return a positive mean ROI. Further fixed cost, would result a
negative ROI.

```{r, eval = TRUE}
#add code here 

validation <- validation %>% 
  mutate(exp_return = (installment*term_months/loan_amnt - 1)*(1-p_val) + (-0.6 * p_val))

validation <- validation %>% 
  mutate(prob_default = p_val)

validation <- validation[order(-validation$exp_return), ]

```

```{r, eval = TRUE}
n_values <- seq(1, nrow(validation), by = 1000) # Change in increments of 10 for efficiency
realized_returns <- numeric(length(n_values))

# Calculate realized return for the top n loans
for (j in 1:length(n_values)) {
  n <- n_values[j]
  ROI <- 0
  
  for (i in 1:n) {
    if (validation[i, ]$default == 0 & validation[i, ]$prob_default <= 16.5) {
      # Non-default: Calculate ROI based on actual return
      ROI <- ROI + ((validation[i, ]$term_months * validation[i, ]$installment) / validation[i, ]$loan_amnt - 1)
    } else if (validation[i, ]$default == 1 & validation[i, ]$prob_default <= 16.5){
      # Default: Apply the -60% loss
      ROI <- ROI - 0.6
    }
  }
  
  # Store the average ROI for the top n loans
  realized_returns[j] <- ROI / n
}
  
df <- data.frame(n = n_values, realized_return = realized_returns)
ggplot(df, aes(x = n, y = realized_return)) +
  geom_line() +
  labs(title = "Realized Return as Portfolio Size (n) Changes",
       x = "Portfolio Size (n)",
       y = "Realized Return")
```

```{r}
profit_700 = 0
for (i in 1:700) {
  if (validation[i, ]$default == 0) {
    profit_700 <- profit_700 + ((validation[i, ]$term_months * validation[i, ]$installment) / validation[i, ]$loan_amnt - 1)
  } else{
    profit_700 <- profit_700 - 0.6
  }
}

# Print the profit for n=700
print(profit_700)
```

For n=700, the profit is **147.1138**. As n increases, the the profit
increases. However, we need to be careful while increasing n, as it
might also start including loans with negative expected_return. The
profit increases until n \< index of negative expected_return.

```{r, eval = TRUE}
#add code here
calculate_profit_per_loss <- function(loss_rate) {
  profit = 0
  for (i in 1:700) {
    if (validation[i, ]$default == 0 & validation[i, ]$prob_default <= 16.5) {
      profit <- profit + ((validation[i, ]$term_months * validation[i, ]$installment) / validation[i, ]$loan_amnt - 1)
    } else if (validation[i, ]$default == 1 & validation[i, ]$prob_default <= 16.5) {
      profit <- profit - loss_rate
    }
  }
  
  return(profit)
}

loss_range <- seq(0.3, 0.9, 0.1)
profit_per_loss_values <- sapply(loss_range, calculate_profit_per_loss)


loss_vs_profit <- tibble(
  loss_rate = seq(0.3, 0.9, 0.1),
  profit = round(profit_per_loss_values, 2)
)

print(loss_vs_profit)
```

The profit reduces, as the loss percentage increases. Hence, the profit
is sensitive and inversely related to loss.

```{r}

# Convert columns to factors and create a binary default character variable
lc_clean <- lc_clean %>%
  mutate(
    def_chr = as.factor(ifelse(default == "1", "D", "P")),  # "D" for default, "P" for paid
    term = factor(term),                                     # Convert 'term' to a factor
    grade = factor(grade),                                   # Convert 'grade' to a factor
    home_ownership = factor(home_ownership),                 # Convert 'home_ownership' to a factor
    emp_length = factor(emp_length)                          # Convert 'emp_length' to a factor
  )

# Create additional feature columns for modeling
lc_clean <- lc_clean %>% 
  mutate(
    income_loan_ratio = annual_inc / loan_amnt,              # Calculate income-to-loan ratio
    log_inc = log(annual_inc + 1),                           # Log-transform of annual income
    term_grade_intr = as.numeric(term) * as.numeric(grade),  # Interaction term between 'term' and 'grade'
    inc_grade_intr = annual_inc * as.numeric(grade)          # Interaction term between income and 'grade'
  )

# Stratified random sampling based on loan 'grade'
set.seed(121)
train_indices <- createDataPartition(lc_clean$grade, p = 0.7, list = FALSE)  # 70% for training
training <- lc_clean[train_indices, ]  # Training set
testing <- lc_clean[-train_indices, ] 

# Define formula for the logistic model
formula <- def_chr ~ annual_inc + loan_amnt + income_loan_ratio + log_inc + term + grade + emp_length + inc_grade_intr + term_grade_intr

# Create model matrix for training data and extract target variable
X_train <- model.matrix(formula, data = training)[, -1]
y_train <- training$def_chr

# Fit LASSO model with cross-validation to select optimal lambda
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1, family = "binomial")

# Extract the best lambda value based on minimum cross-validation error
min_lambda <- lasso_model$lambda.min

# Fit the LASSO model with the optimal lambda
model_lasso <- glmnet(X_train, y_train, family = "binomial", alpha = 1, lambda = min_lambda)

# Predict on the training set and calculate training AUC
train_pred <- predict(model_lasso, newx = X_train, type = "response")
train_roc <- roc(y_train, as.vector(train_pred))
train_auc <- auc(train_roc)

# Prepare testing data and calculate AUC for validation
X_val <- model.matrix(formula, data = testing)[, -1]
y_val <- as.numeric(as.character(testing$def_chr))

# Predict on the validation set and calculate validation AUC
val_pred <- predict(model_lasso, newx = X_val, type = "response")
val_roc <- roc(y_train, as.vector(train_pred))
val_auc <- auc(val_roc)

# Display AUC for training and testing sets
cat("AUC for LASSO model testing set:", val_auc, "\n")
cat("AUC for LASSO model training set:", train_auc, "\n")




```

```{r}
# Read the assessment data CSV file and create new features
lc_assessment <- read_csv("D:\\data_science_business\\assignment2\\Assessment Data_2024.csv") %>%
  mutate(
    term = as.factor(`term (months)`),                       # Convert 'term' to factor
    grade = as.factor(grade),                                # Convert 'grade' to factor
    home_ownership = factor(home_ownership),                 # Convert 'home_ownership' to factor
    emp_length = factor(emp_length),                         # Convert 'emp_length' to factor
    income_loan_ratio = annual_inc / loan_amnt,              # Calculate income-to-loan ratio
    log_inc = log(annual_inc + 1),                           # Apply log transformation to income
    term_grade_intr = as.numeric(term) * as.numeric(grade),  # Interaction term: 'term' * 'grade'
    inc_grade_intr = annual_inc * as.numeric(grade)          # Interaction term: income * 'grade'
  )

# Align factor levels in assessment data with training data for consistency
lc_assessment$term <- factor(lc_assessment$term, levels = levels(training$term))
lc_assessment$grade <- factor(lc_assessment$grade, levels = levels(training$grade))
lc_assessment$home_ownership <- factor(lc_assessment$home_ownership, levels = levels(training$home_ownership))
lc_assessment$emp_length <- factor(lc_assessment$emp_length, levels = levels(training$emp_length))

# Define the model formula for prediction without the response variable
predict_formula <- ~ annual_inc + loan_amnt + income_loan_ratio + log_inc + term + grade + emp_length + inc_grade_intr + term_grade_intr

# Create a model matrix for assessment data to match the structure of training data
X_asmnt <- model.matrix(predict_formula, data = lc_assessment)[, -1]

# Identify missing columns in X_assessment and add them as zero columns for consistency
missing_cols <- setdiff(colnames(X_train), colnames(X_asmnt))
if (length(missing_cols) > 0) {
  X_asmnt <- cbind(X_asmnt, Matrix(0, nrow = nrow(X_asmnt), ncol = length(missing_cols)))
  colnames(X_asmnt)[(ncol(X_asmnt) - length(missing_cols) + 1):ncol(X_asmnt)] <- missing_cols
}

# Ensure column order in X_assessment matches that in X_train
X_asmnt <- X_asmnt[, colnames(X_train)]

# Predict default probabilities using the trained LASSO model
predict_probs <- predict(model_lasso, newx = X_asmnt, type = "response")

# Calculate expected returns for each loan in the assessment data
lc_assessment <- lc_assessment %>%
  mutate(
    predicted_prob = as.vector(predict_probs),               # Store predicted probabilities
    installment = as.numeric(installment),                   # Convert 'installment' to numeric
    term_months = as.numeric(`term (months)`),               # Convert 'term' to numeric
    loan_amnt = as.numeric(loan_amnt),                       # Convert 'loan_amnt' to numeric
    return_if_paid = (installment * term_months) / loan_amnt - 1, # Expected return if loan is paid in full
    return_if_default = -0.6,                                # Expected return if loan defaults
    expected_return = return_if_paid * (1 - predict_probs) + return_if_default * predict_probs # Calculate expected return
  )

# Select the top 240 loans based on expected return
selected_loans <- lc_assessment %>%
  arrange(desc(expected_return)) %>%                         # Sort loans by expected return
  slice_head(n = 240) %>%                                    # Select top 240 loans
  mutate(roopa_yashaswini = 1) %>%                           # Label selected loans with 1
  select(`Loan number`, roopa_yashaswini)

# Label remaining (non-selected) loans with 0
non_selected_loans <- lc_assessment %>%
  filter(!`Loan number` %in% selected_loans$`Loan number`) %>% # Filter non-selected loans
  mutate(roopa_yashaswini = 0) %>%                           # Label non-selected loans with 0
  select(`Loan number`, roopa_yashaswini)

# Combine selected and non-selected loans, keeping original order
final_selection <- bind_rows(selected_loans, non_selected_loans) %>%
  arrange(`Loan number`) %>%                                 # Arrange by loan number
  rename(loan_number = `Loan number`)                        # Rename column for output

# Save the final selection to a CSV file
write.csv(final_selection, "Roopa_Yashaswini.csv", row.names = FALSE)

```

```{r, eval = TRUE}
lc_assessment<- read_csv("Assessment Data_2024.csv") %>%  #load the data 
  clean_names() %>% # use janitor::clean_names() 
  mutate(
    issue_d = mdy(issue_d),  # lubridate::mdy() to fix date format
    term = factor(term_months),     # turn 'term' into a categorical variable
    delinq_2yrs = factor(delinq_2yrs)) # turn 'delinq_2yrs' into a categorical variable
```
